---
output: 
  github_document
bibliography: inst\\REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  eval = FALSE
)
```

# ir

ir is an R package that contains simple functions to import, handle and preprocess infrared spectra. Infrared spectra are stored as list columns in `data.frame`s to enable efficient storage of metadata along with the spectra and support further analyses containing other data for the same samples.

Supported file formats for import currently are:

1. .csv files with indiidual spectra.
2. Thermo Galactics .spc files with individual spectra.

Provided functions for preprocessing and general handling are:

1. baseline correction with:
    * a polynomial 
    * a convex hull procedure.
2. binning.
3. clipping.
4. interpolating (resampling, linearly).
5. replacing selected parts of a spectrum by a straight line.
6. averaging spectra within specified groups.
7. normalise spectra:
    * to the maximum intensity
    * to the intensity at a specific x value
    * so that all itensity values sum to 1.
8. smoothing: 
    * Savitzky-Golay smoothing 
    * Fourier smoothing.

### How to install

You can install ir from GitHub using R via:
```{r installation, eval = FALSE}
remotes::install_github(repo = "henningte/ir")
```

### How to use

You can load ir in R with:
```{r load_ir}
library(ir)

# load additional packages needed for this tutorial
library(ggplot2)
library(magrittr)
```

You can load the sample data with:
```{r sample_data_load}
ir::sample_data
```

`sample_data` is an object of class `ir`. An Object of class `ir` is basically a `data.frame` where each row represents one infrared measurement and column `spectra` contains the infrared spectra (one per row) and columns `measurement_id` and  `sample_id` represent identifiers for each measurement and sample, respectively.
```{r sample_data_inspect}
head(sample_data, 2)
```

This allows effectively storing repeated measurements for the same sample in the same table, as-well-as any metadata and accessory data (e.g. nitrogen content of the sample).

The column `spectra` is a list column of `data.frame`s, meaning that each cell in `sample_data` contains for column `spectra` a `data.frame`. For example, the first element of `sample_data$spectra` represents the first spectrum as a `data.frame`
```{r sample_data_inspect_spectra}
ir::ir_get_spectrum(sample_data, what = 1)[[1]] %>% 
  head(10)
```

Column `x` represents the x values (in this case wavenumbers [cm<sup>-1</sup>]) and column `y` the corresponding intensity values.

A simple workflow would be, for example, to baseline correct the spectra, then bin them to bins with a width of 10 wavenumber units, then normalise them so that the maximum intensity value is 1 and the minimum intensity value is 0 and then plot the baseline corrected samples for each sample:
```{r sample_data_workflow}
ir::ir_bc(sample_data, method = "rubberband") %>%    # baseline correction
  ir::ir_bin(width = 10) %>%                         # binning
  ir::ir_normalise(method = "zeroone") %>%           # normalisation
  plot() + ggplot2::facet_wrap(~ sample_id)          # plot
```

### How to cite

Please cite this R package as:

> Henning Teickner, (`r format(Sys.Date(), "%Y")`). _ir: A Simple Package to Handle and Preprocess Infrared Spectra'_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://github.com/henningte/ir>.


### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### Sources

This packages was developed in R (`r R.Version()$version.string`) [@RCoreTeam.2019] using functions from devtools [@Wickham.2019], usethis [@Wickham.2019b] and rrtools [@Marwick.2019].

### References
